---
title: "Visualising Safety data"
author: "Rob Solomon"
date: "2023-06-30"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    number_sections: true
    df_print: paged

---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,warning = FALSE)
```

# Visualisation of Safety Data Using R-Markdown {.tabset}

## Initial loading and opening of libraries

```{r}
# Install readxl R package - do this only once
#install.packages("readxl")

# open libraries
library(readxl)
library(tidyverse)
library(data.table)
library(ggplot2)
```

## Reading Data {.tabset}

### Introduction

Here I need to give ambit of an introduction on reading and initial wrangling

### Get the Data

```{r warning = FALSE,error=FALSE}

safety_data <- read_excel("safety data - Dr Apurna Ghosh.xlsx")
```

### Basic Data Wrangling

Read the Convert text format to date format and create two new columns using mutate.

```{r}
safety_data <- safety_data %>% 
  mutate(
    Date =as.Date(Date,"%d/%m/%Y"),
    month=month(Date),
         year=year(Date))
```

### Quick Check on Numbers

```{r}

#
# quick look to see if employee numbers are unique
#

multiple <- safety_data %>%
  group_by(EmployeeID) %>%
  summarise(n = n()) %>%
  filter(n > 1)

print(paste0("There are ",nrow(multiple)," EmployeeID's with multiple events"))


```

## Visualisation {.tabset}

### Run Charts{.tabset}

#### Introduction

Wrangle the data
Create a new column of month start dates using as.Date() and paste0() functions
group the data in groups of months based on month_start column
produce a new column n_events containing the total number of events using n().
other common sumarising functions are mean(x , na.rm=TRUE/FALSE), sd(x , na.rm=TRUE/FALSE) -Standard deviation, max(x , na.rm=TRUE/FALSE),
min(x , na.rm=TRUE/FALSE),quantile(x,prob=0.9)

#### Data Wrangling

```{r}


safety_data <- safety_data %>% 
  mutate(month_start=as.Date(paste0(year,"-",month,"-",01)))
trend <- safety_data %>% 
  group_by(month_start) %>% 
  summarise(n_events=n()) 
```

#### Run Chart {.tabset}

##### Basic Run Chart

```{r}
# create a basic plot variable using ggplot
p <- trend %>% ggplot(aes(x=month_start,y=n_events))+
  geom_point()
# Display it
plot(p)
```

##### Add titles

```{r}
# improve the chart with some labels
p <- p +
  labs(title = "Trend of Monthly Events",
       subtitle = "all events",
       x="Date",
       y="Number of Events")
plot(p)
```

##### Add a Trend line

```{r}

p1 <- p+
  geom_smooth()
plot(p1)
```

##### Change Degree of smoothing

```{r}
# you can alter the degree of smoothing using the span parameter
p1 <- p+
  geom_smooth(span=.08)
plot(p1)
```

##### Change the look of the chart {.tabset}

###### Change colour of line

```{r}

p1 <- p+
  geom_smooth(span=.08,colour="blue")
plot(p1)
```



##### remove error region

```{r}

p1 <- p+
  geom_smooth(span=.08,colour="blue",se=FALSE)
plot(p1)
```



##### Change the colour and intensity of shading

```{r}
# Change the colour and intensity of shading
p1 <- p+
  geom_smooth(span=.08,colour="blue",fill="blue",alpha=0.5)
plot(p1)
```



##### Change the colour and intensity of shading

```{r}
p1 <- p+
  geom_smooth(span=.08,colour="blue",fill="blue",alpha=0.3)
plot(p1)
```

#### Rolling averages

```{r}
# This is a great visualisation method as the trend is a central estimate, many trending processes such as 
# rolling averages tend to lag the data lets add a typical 12 month rolling mean to the plot.
# to do this we need to create an additional column and install the "zoo" package for rolling statistics
#install.packages("zoo")

# lets now add a 3 month moving average to the data
library(zoo)
trend <- trend %>% 
  mutate(moving_average=rollmean(n_events,k=3,fill=NA))

p2 <- p1 +
  geom_line(data=trend,aes(x=month_start,y=moving_average),colour="red")
p2
```

#### Multiple lines on the same chart {.tabset}

##### Using facet wrap

```{r}
# now plot multiple views on the same chart 
trend <- safety_data %>% 
  mutate(month_start=as.Date(paste0(year,"-",month,"-",01))) %>% 
  group_by(month_start,FullOrPartTime) %>% 
  summarise(n_events=n()) 
p <- trend %>% ggplot(aes(x=month_start,y=n_events,colour=FullOrPartTime,fill=FullOrPartTime))+
  geom_point()
p <- p +
  labs(title = "Trend of Monthly Events",
       subtitle = "all events",
       x="Date",
       y="Number of Events")

print(p)

```

##### Adding trend lines

```{r}

p1 <- p +
  geom_smooth()
print(p1)
```


```{r}
# # now plot multiple views on the same chart 
# trend <- safety_data %>% 
#   mutate(month_start=as.Date(paste0(year,"-",month,"-",01))) %>% 
#   group_by(month_start,BodyPart) %>% 
#   summarise(n_events=n()) 
# p <- trend %>% ggplot(aes(x=month_start,y=n_events,colour=BodyPart,fill=BodyPart))+
#   geom_point()
# p <- p +
#   labs(title = "Trend of Monthly Events",
#        subtitle = "all events",
#        x="Date",
#        y="Number of Events")
# print(p)
```

##### A more complex example {.tabset}

Produce a series of plots on the trends of the top 10 body part locations

###### Data Wrangling

```{r}
# First create a data frame containing a list of body parts and frequency of incidents
# using group and summarise.
# Arrange in decreasing order
# and take a slice of the top 10
top_10_body_parts <- safety_data %>% 
  mutate(BodyPart=as.character(BodyPart)) %>% 
  group_by(BodyPart) %>% 
  summarise(n=n()) %>% 
  arrange(-n) %>% 
  slice(1:10)

# Now right join this with the original safety data set to obtain a data from containing 
# all the data for the to 10 body part locations

top_10_body_part_data <- right_join(safety_data,top_10_body_parts)
# now generate the monthly data for each data set
top_10_body_part_trend <- top_10_body_part_data %>% 
  group_by(month_start,BodyPart) %>% 
  summarise(n_events=n())
```

###### Output from Wrangling

```{r}

top_10_body_part_trend

```
###### Graphical output {.tabset}


####### Graphical output

```{r}
p <- top_10_body_part_trend %>% ggplot(aes(x=month_start,y=n_events,colour=BodyPart,fill=BodyPart))+
  geom_line()
p <- p +
  labs(title = "Trend of Monthly Events",
       subtitle = "Grouped by body part",
       x="Date",
       y="Number of Events")
print(p)
```

####### Add a trend

```{r}
p1 <- p +
  geom_smooth()
print(p1)
```

####### facet wrap

```{r}
p <- top_10_body_part_trend %>% 
  ggplot(aes(x=month_start,y=n_events,colour=BodyPart)) +
  geom_point() +
  facet_wrap(facets = vars(BodyPart))
```

####### with a trend

```{r}
print(p)
p <- top_10_body_part_trend %>% 
  ggplot(aes(x=month_start,y=n_events,colour=BodyPart)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(facets = vars(BodyPart),scales = "fixed")
print(p)
```


### Pareto Charts


```{r}
## Pareto

top_10_body_part_data$BodyPart <- factor(top_10_body_part_data$BodyPart,
                                         levels=top_10_body_parts$BodyPart)

top_10_body_parts$BodyPart <-
  factor(top_10_body_parts$BodyPart ,
         levels = top_10_body_parts$BodyPart [order(-top_10_body_parts$n)])


p <- top_10_body_parts %>% ggplot(aes(x=BodyPart,y=n,fill=BodyPart))+
  geom_col(alpha=0.7)
print(p)
p <- p+
  theme(axis.text.x = element_text(
    # rotates the x axis labels
    angle = 90,
    hjust = 1,
    vjust = 0.5
  )) 
print(p)
```


```{r}
top_10_body_part_annual_trend <- top_10_body_part_data %>% 
  group_by(year,BodyPart) %>% 
  summarise(n_events=n()) 

p <- top_10_body_part_annual_trend %>%
  ggplot(aes(x = BodyPart, y = n_events, fill = BodyPart)) +
  geom_col(alpha = 1) +
  theme(legend.position = "none",
        axis.text.x = element_text(
          # rotates the x axis labels
          angle = 90,
          hjust = 1,
          vjust = 0.5
        )) +
  facet_wrap(facets = vars(year))
p

```

### Pie Charts


```{r}

library(scales)
tot_events <- sum(top_10_body_parts$n)
top_10_body_parts$frac <- top_10_body_parts$n/tot_events
p <- top_10_body_parts %>%
  ggplot(aes(x = "", y = frac,fill=BodyPart)) +
  geom_col(alpha=0.5)+
   geom_text(aes(label = paste(round(frac*100,1),"%")),
            position = position_stack(vjust = .5),
            size=3,
            angle=45
            )+
  coord_polar("y", start=0)+
  theme_void()

p
```
